---
title: "ROS - causal inference and randomised experiments notes"
author: "Erika Duan"
date: "`r Sys.Date()`"

execute:
  echo: true
  output: false
  message: false
  warning: false

format:
  gfm:
    toc: true
    html-math-method: webtex
---

```{r}
#| echo: false 

# Load required R packages -----------------------------------------------------
if (!require("pacman")) install.packages("pacman")
pacman::p_load() 
```

# Basics of causal inference  

Causal inference is a common concept in clinical trials. These trials generally aim to answer the questions:   

+ What would have happened if person A received treatment A compared to treatment B or a placebo?   
+ Do we have clear evidence to change clinical practice (to improve patient outcomes)?    

A useful clinical trial clearly identifies whether an **average causal effect** exists between different treatments or a treatment versus placebo.   

We can illustrate the basics of causal inference through an example using synthetic data.    

```{r}
# Scenario 1 -------------------------------------------------------------------
# Question: do omega-3 fish oil supplements impact blood pressure?  

# n = 8 
# n = 4 maintaining current diets and additional fish oil supplement consumption
# n = 4 maintaining current diets without fish oil supplements  
# We assume all study participants complete the treatment that they are assigned


```

